The function--sheaf dictionary for smooth commutative group schemes over finite fields, with applications over local fields

### 2014-08-22 dictionary.tex

Returning to the case of connected commutative algebraic groups $G$ over $\FF_q$, we confirm that, in this case, the category of quasicharacter sheaves on $G$ is equivalent to the category of characters $\psi$ of $G(\FF_q)$, the equivalence being given by pullback by the Lang isogeny for $G$. 
%Indeed, one may show that isomorphism classes of quasicharacter sheaves on $G$ are determined by classes of extensions of $\bG$ by $\EEx$ which are fixed by Frobenius, given by $\Hh^2(\bG,\EEx)^{\Gal({\bar\FF}_q/\FF_q)}$.
%
%The appearance of $\Hh^2(\bG,\EEx)^{\Gal({\bar\FF}_q/\FF_q)}$ in these two cases is interesting: on the one hand, when $G$ is a connected commutative algebraic group over $\Fq$, isomorphism classes of quasicharacter sheaves on $G$ are measured by $\Hh^2(\bG,\EEx)^{\Gal({\bar\FF}_q/\FF_q)}$ and the trace of Frobenius gives an isomorphism $T_G: \QCiso{G} \to \Hom(G(\FF_q),\EEx)$; on the other hand, if $G$ an étale commutative group scheme over $\FF_q$, then $\Hh^2(\bG,\EEx)^{\Gal({\bar\FF}_q/\FF_q)}$ measures the kernel of $T_G: \QCiso{G} \to \Hom(G(\FF_q),\EEx)$. 
%This observation makes plain the fact that w

As we show in the paper, while characters of connected commutative algebraic groups $G$ can be geometrized using  local systems trivialized by discrete isogenies to $G$, this is not the right perspective if one wishes to geometrize characters of smooth commutative group schemes, such as étale commutative group schemes. Indeed, quasicharacter sheaves on étale commutative group schemes that are trivialized by discrete isogenies are invisible to the trace of Frobenius.

@@

We close the introduction with an overview of the structure of this paper.
After setting some basic notation in Section~\ref{sec:notation},
in Section~\ref{sec:category} we define the category of quasicharacter sheaves.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism $\TrFrob{G} : \QCiso{G} \to \Hom(G(k), \EEx)$.
We also define pullbacks of quasicharacter sheaves along morphisms of group schemes.  Applying pullbacks to the component group
sequence for $G$ yields the diagram
 \begin{equation} \label{eq:intro-snake}
  \begin{tikzcd}[row sep=20]
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar & G(\Fq)^* \rar & G^0(\Fq)^* \rar & 0,\\
  \end{tikzcd}
  \end{equation}
which is central to the proof of Theorem \ref{thm:snake}.  Much of Sections \ref{sec:etale}, \ref{sec:disc-isog} and \ref{sec:main}
are devoted to proving that the rows of \eqref{eq:intro-snake} are exact and determining the kernels and cokernels of the vertical maps.

Section~\ref{sec:etale} is devoted to the left-most vertical morphism in \eqref{eq:intro-snake}
In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{sec:descentG} and \ref{sec:finite} describe bounded quasicharacter sheaves and finite quasicharacter sheaves, respectively,
forming two important subcategories of $\QC(G)$.
%These subcategories play a key role in the proof of Proposition~\ref{prop:restriction}.
In Section \ref{sec:connected} we show that all quasicharacter sheaves on $G$ are finite quasicharacter sheaves when $G$ is connected, and that Trace of Frobenius is an isomorphism in this case.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and find a cohomological interpretation
of quasicharacter sheaves in this case.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
%Section~\ref{sec:snake} presents the main theorem of this paper: Trace of Frobenius
%is a functorial isomorphism from isomorphism classes of flabby quasicharacter sheaves on $G$
%to characters of $G(\Fq)$.

@@

\subsection{Objects, when $G$ is étale}\label{sec:SandT}

@@

Consider the short exact sequence
defining the component group scheme for $G$:
\begin{equation}\label{eq:pi0}
\begin{tikzcd}
0 \rar & G^0 \arrow{r}{\iota_0} & G \arrow{r}{\pi_0} & \pi_0(G) \rar & 0.
\end{tikzcd}
\end{equation}
Since $\pi_0(G)$ is a étale commutative group scheme -- and thus smooth --
Proposition~\ref{prop:pullback} implies that \eqref{eq:pi0} defines the sequence of functors
\begin{equation}\label{eq:pi1}
\begin{tikzcd}
\QC(0) \rar & \QC(\pi_0(G)) \arrow{r}{\pi_0^*} & \QC(G) \arrow{r}{\iota_0^*} & \QC(G^0) \rar & \QC(0)
\end{tikzcd}
\end{equation}
and therefore, after passing to isomorphism classes, the sequence of abelian groups
\begin{equation}\label{eq:pi2}
\begin{tikzcd}
0 \rar &
\QCiso{\pi_0(G)} \arrow{r}{\pi_0^*} & \QCiso{G} \arrow{r}{\iota_0^*} & \QCiso{G^0} \rar & 0.
\end{tikzcd}
\end{equation}
 Note that we found the groups $\QCiso{\pi_0(G)}$ and $\QCiso{G^0}$
in Sections~\ref{sec:SandT} and \ref{sec:connected}, respectively.
We will shortly see that \eqref{eq:pi2} is exact.

>>>>

Returning to the case of connected commutative algebraic groups $G$ over $\FF_q$, we confirm that, in this case, the category of quasicharacter sheaves on $G$ is equivalent to the category of characters $\psi$ of $G(\FF_q)$, the equivalence being given by pullback by the Lang isogeny for $G$. 
%
%The appearance of $\Hh^2(\bG,\EEx)^{\Gal({\bar\FF}_q/\FF_q)}$ in these two cases is interesting: on the one hand, when $G$ is a connected commutative algebraic group over $\Fq$, isomorphism classes of quasicharacter sheaves on $G$ are measured by $\Hh^2(\bG,\EEx)^{\Gal({\bar\FF}_q/\FF_q)}$ and the trace of Frobenius gives an isomorphism $T_G: \QCiso{G} \to \Hom(G(\FF_q),\EEx)$; on the other hand, if $G$ an etale commutative group scheme over $\FF_q$, then $\Hh^2(\bG,\EEx)^{\Gal({\bar\FF}_q/\FF_q)}$ measures the kernel of $T_G: \QCiso{G} \to \Hom(G(\FF_q),\EEx)$. 
%This observation makes plain the fact that  . . .
%
As we show in the paper, while characters of connected commutative algebraic groups $G$ can be geometrized using  local systems trivialized by discrete isogenies to $G$, this is not the right perspective if one wishes to geometrize characters of smooth commutative group schemes, such as étale commutative group schemes. Indeed, quasicharacter sheaves on étale commutative group schemes that are trivialized by discrete isogenies are invisible to the trace of Frobenius.

@@

We close the introduction with an overview of the structure of this paper.
After setting some basic notation in Section~\ref{sec:notation},
in Section~\ref{sec:category} we define the category of quasicharacter sheaves.
Section~\ref{sec:Frob} introduces the trace of Frobenius homomorphism $\TrFrob{G} : \QCiso{G} \to \Hom(G(k), \EEx)$.
We also define pullbacks of quasicharacter sheaves along morphisms of group schemes.  Applying pullbacks to the component group
sequence for $G$ yields the diagram
 \begin{equation} \label{eq:intro-snake}
  \begin{tikzcd}[row sep=20]
    0 \rar & \QCiso{\pi_0(G)} \rar \dar{\TrFrob{\pi_0(G)}}
    & \QCiso{G} \rar \dar{\TrFrob{G}} & \QCiso{G^0} \rar \dar{\TrFrob{G^0}} & 0\\
    0 \rar & \pi_0(G)(\Fq)^* \rar & G(\Fq)^* \rar & G^0(\Fq)^* \rar & 0,\\
  \end{tikzcd}
  \end{equation}
which is central to the proof of Theorem~\ref{thm:snake}.  Much of Sections~\ref{sec:etale}, \ref{sec:disc-isog} and \ref{sec:main}
are devoted to proving that the rows of \eqref{eq:intro-snake} are exact and determining the kernels and cokernels of the vertical maps.

Section~\ref{sec:etale} is devoted to the left-most vertical morphism in \eqref{eq:intro-snake}: studying $\TrFrob{G}$ for $G$ \'etale.  We begin by giving a reinterpretation of
$\QC(G)$ in this case using stalks.  After a brief review of the Hochschild-Serre spectral sequence in Section~\ref{ssec:E},
we define in Section~\ref{ssec:S} an isomorphism $S_G$ from $\QCiso{G}$ to the second cohomology of the total space of the spectral sequence
\[
E_2^{p,q} \ceq \Hh^p(\Weil{}, \Hh^q(\bG, \EEx)) \Rightarrow \Hh^{p+q}(\Weil{} \ltimes \bG, \EEx).
\]
Paired with the short exact sequence
\[
  0 \to \Hh^0(\Weil{},\Hh^2({\bar G},\EEx)) \to \Hh^2(E^\bullet_G) \to \Hh^1(\Weil{},\Hh^1({\bar G},\EEx)) \to 0
\]
arising from the spectral sequence, the isomorphism $S_G$ allows us to describe the kernel and cokernel of $\TrFrob{G}$ in Section \ref{ssec:SandT}.
We close our discussion of \'etale $G$ with a description, in Section~\ref{ssec:mor-etale}, of $\Hom(\qcs{L},\qcs{L}')$ for $\qcs{L}, \qcs{L}' \in \QC(G)$.

Section~\ref{sec:disc-isog} focuses on the right-hand side of \eqref{eq:intro-snake},
giving descriptions of quasicharacter sheaves that are particularly useful when $G$ is connected.  In Section \ref{ssec:descentG}
we consider quasicharacter sheaves that arise via base change to $\bFq$ from local systems on $G$, and in Section \ref{ssec:discrete_isogenies}
those that arise via pushforward from a constant sheaf along a discrete isogeny $H \to G$.  When $G$ is connected,
we show in Section~\ref{} that every quasicharacter sheaf can be described in both of these ways.  We also show that $\TrFrob{G}$ is an isomorphism
in this case.

In Section~\ref{sec:basechange} we study how quasicharacter sheaves behave under base change
and Weil restriction.
Sections~\ref{ssec:descentG} and \ref{sec:finite} describe bounded quasicharacter sheaves and finite quasicharacter sheaves, respectively,
forming two important subcategories of $\QC(G)$.
%These subcategories play a key role in the proof of Proposition~\ref{prop:restriction}.
In Section \ref{ssec:connected} we show that all quasicharacter sheaves on $G$ are finite quasicharacter sheaves when $G$ is connected, and that Trace of Frobenius is an isomorphism in this case.
We shift to \'etale group schemes $G$ in Section~\ref{sec:etale} and find a cohomological interpretation
of quasicharacter sheaves in this case.
In Section~\ref{sec:restriction} we show that for non-connected $G$, restriction to the identity component
induces an essentially surjective functor of quasicharacter sheaves.
%Section~\ref{sec:snake} presents the main theorem of this paper: Trace of Frobenius
%is a functorial isomorphism from isomorphism classes of flabby quasicharacter sheaves on $G$
%to characters of $G(\Fq)$.

@@

\subsection{Objects, when $G$ is etale}\label{ssec:SandT}

@@

Consider the short exact sequence
defining the component group scheme for $G$:
\begin{equation}\label{eq:pi0}
\begin{tikzcd}
0 \rar & G^0 \arrow{r}{\iota_0} & G \arrow{r}{\pi_0} & \pi_0(G) \rar & 0.
\end{tikzcd}
\end{equation}
Since $\pi_0(G)$ is a étale commutative group scheme -- and thus smooth --
Proposition~\ref{prop:pullback} implies that \eqref{eq:pi0} defines the sequence of functors
\begin{equation}\label{eq:pi1}
\begin{tikzcd}
\QC(0) \rar & \QC(\pi_0(G)) \arrow{r}{\pi_0^*} & \QC(G) \arrow{r}{\iota_0^*} & \QC(G^0) \rar & \QC(0)
\end{tikzcd}
\end{equation}
and therefore, after passing to isomorphism classes, the sequence of abelian groups
\begin{equation}\label{eq:pi2}
\begin{tikzcd}
0 \rar &
\QCiso{\pi_0(G)} \arrow{r}{\pi_0^*} & \QCiso{G} \arrow{r}{\iota_0^*} & \QCiso{G^0} \rar & 0.
\end{tikzcd}
\end{equation}
 Note that we found the groups $\QCiso{\pi_0(G)}$ and $\QCiso{G^0}$
in Sections~\ref{ssec:SandT} and \ref{ssec:connected}, respectively.
We will shortly see that \eqref{eq:pi2} is exact.
